\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-12-02 doc/publications/plos16/plos16.tex
added reference to plos16.tex

Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs, termed
single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
separately~\cite{CCC2007}.

It has since been
demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
is inadequate~\cite{Manolio2009,Yang2011}.

A tree based model is capable of taking this interaction of variables
into account. In addition, random forests are well suited for
processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
Secondly, random forests are also very easy to parallelize. As the forest is a sum of decision trees, it it possible to grow
separate trees on different processors and combine the trees. See the
discussion of the properties of random forests in Section \ref{section:methods}.

@@

Since the original implementation of random forests, there have been a number of developments leading to a better understanding of
the algorithm. 
\begin{itemize}
\item random forests are resistant to overfitting. However it is not true that they will not overfit,
  ~\cite{Segal.2004} gives examples of overfitting in cases where:
  \begin{itemize}
  \item a single tree is the most appropriate model; and
  \item the variables are highly correlated so that the bootstrap resampling does not give rise to uncorrelated trees.
  \end{itemize}
However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.

\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection via two mechanisms:
  \begin{enumerate}
  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
    bias. A uninformative variable with a large number of levels may be selected over a more informative variable with
    fewer levels;
  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$ of the data without replacement seems to be an
    effective strategy for removing the problem.
  \end{enumerate}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \ntree.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 4: Sensitivity 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scalability analysis}

In this section we explore the performance of \cursedforest\ by testing its ability to scale to different sizes of data
and computational resources.

@@

\section{Conclusion}

We have demonstrated that using a different model of paralleization can extend the random forest to delaing with the
case of an extremely large number of variables. By comparing this implementation to other implentations (including those
optimized for large data sets) we have demonstrated the utility of this implentation.

>>>>

Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs,
termed single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP
separately~\cite{CCC2007}.  It has since been demonstrated that not taking joint interactions between SNPs into account
when filtering for potential disease mutations is inadequate~\cite{Manolio2009,Yang2011}.

A tree based model is capable of taking this interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelize. As the forest is a sum of decision trees, it it possible to grow separate
trees on different processors and combine the trees. See the discussion of the properties of random forests in Section
\ref{section:methods}.

@@

Since the original implementation of random forests, there have been a number of developments leading to a better
understanding of the algorithm.
\begin{itemize}
\item random forests are resistant to overfitting. However it is not true that they will not overfit,
  ~\cite{Segal.2004} gives examples of overfitting in cases where:
  \begin{itemize}
  \item a single tree is the most appropriate model; and
  \item the variables are highly correlated so that the bootstrap resampling does not give rise to uncorrelated trees.
  \end{itemize}
However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.
\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection
  via two mechanisms:
  \begin{enumerate}
  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
    bias. A uninformative variable with a large number of levels may be selected over a more informative variable with
    fewer levels;
  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$ of the data without replacement seems to be an
    effective strategy for removing the problem.
  \end{enumerate}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
% In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in parallel
% results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
% Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
% run, the ranking of the variables will stabilise with growing \ntree.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 4: Sensitivity 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scalability analysis}
In this section we explore the performance of \cursedforest\ by testing its ability to scale to different sizes of data
and computational resources.

@@

\section{Conclusion}
We have demonstrated that using a different model of paralleization can extend the random forest to delaing with the
case of an extremely large number of variables. By comparing this implementation to other implentations (including those
optimized for large data sets) we have demonstrated the utility of this implentation.
